---
title: "Predicting succesful execution of dumbbell excercises from accelerometers data."
author: "E. Donis"
date: "10/17/2020"
output: html_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80),tidy = TRUE)
library(caret); library(tidyverse)

```
## Overview
A classification model was built to predict from accelerometers data the manner in which one does dumbbell exercises. The selected model is a Random Forest which predicted correctly 95% of the given 20 test samples. This report captures the process taken in building such model. 

## Data
The training data was obtained and loaded into R as follow:
```{r, message=FALSE, cache=TRUE}
train_link <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
download.file(train_link,destfile = "./data/pml-train.csv",method = "curl")
ds.train <- read.csv("./data/pml-train.csv")
dim(ds.train)
```
The outcome variable is labeled `classe` and the distribution is fairly balance with Class A having more observations than the rest. 
```{r, message=FALSE,fig.height=2.3,fig.width=4,fig.align='center'}
qplot(x=classe, fill= classe,data = ds.train, ylab = "count", main="Distribution of outcome")
```
*A - exactly according to the specification  
*B - throwing the elbows to the front  
*C - lifting the dumbbell only halfway  
*D - lowering the dumbbell only halfway 
*E - throwing the hips to the front

### Pre-processing
In the raw dataset, accelerometers data is available for 4 different locations i.e. on the belt, forearm, arm, and dumbbell of the 6 participants that took place in the controlled experiment. For each of the location, there are 38 corresponding variables e.g.   
```{r}
grep("belt",names(ds.train),value = TRUE)
```
More information is available from the website here: <http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har> (see the section on the Weight Lifting Exercise Dataset).

The predictors variables in the training dataset has a considerable amount of empty values
```{r}
sum(is.na(ds.train[,-160]))
```
hence the variables considered for building this model are the direct measurements from the accelerometers in each plane direction (x, y, z).
```{r}
accel <- grep("^accel",names(ds.train),value = TRUE);myvar <- c(accel,"classe")
accel
```
The Principal Component Analysis was applied to the remaining 12 predictor variables, resulting in 8 components needed to capture the 95% of the variance.
```{r}
prePro <- preProcess(ds.train[,accel], method = "pca");prePro
```

## Building Models
Two different models were built and evaluated.  

### Descision tree 'rpart'
The pre-processed PCA components were used to predict a new train dataset to build the `rpart` model on. The model executed 25 Bootstrap repetitions to tune the model.  
```{r, cache=TRUE, message=FALSE}
trainPC <- predict(prePro,ds.train[,accel])
rpart.fit <- train(classe~.,method = "rpart", data = data.frame(trainPC,classe = ds.train$classe))
rpart.fit
```

### Random Forest 'rf'
In this case PCA pre-processing was defined within the `train()` function. Again, sampling took the form of 25 Bootstrap repetitions and 500 trees were built to tune the model.  
```{r,cache=TRUE, warning=FALSE,message=FALSE}
rf.fit <- train(classe~.,method = "rf", data = ds.train[,myvar], preProcess="pca")
rf.fit
```

## Model Selection  

### Evaluating Performance
`rpart` did not yield favorable results with 40% accuracy and not able to predict Classes C and E.
```{r}
table(ds.train$classe,predict(rpart.fit,trainPC))
```
`rf` ,on the other hand, performs much better with 89% accuracy. It can also be noted that it performs the best at predicting the Class A which might be a signal of imbalance given the distribution of the outcome in the train dataset.
```{r}
rf.fit$finalModel
```

### Testing / Predicting
The final validation of the `rf` model was done against the given test data, which was downloaded and loaded into R as follow:
```{r, message=FALSE, cache=TRUE}
test_link <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(test_link,destfile = "./data/pml-testing.csv")
ds.test <- read.csv("./data/pml-testing.csv")
dim(ds.test)
```
The test dataset does not contain the outcome variable and instead, contains a numeric variable, `id_problem`, corresponding to the Prediction Quiz in the Coursera site.
```{r}
range(ds.test[,160])
```
Here is the final prediction from the model which resulted in a 95% score in the prediction Quiz, only missing one of the 20 samples.
```{r}
test.pred <- predict(rf.fit,ds.test[,accel])
t(data.frame(Test_No = ds.test[160],Prediction =test.pred))
```



